---
title: "Shapers and renderers: how blockr.viz is organized"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shapers and renderers: how blockr.viz is organized}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

blockr.viz is the **render layer** for blockr dashboards. It draws rectangular
data frames; it does not compute or reshape them. This vignette explains the
small set of roles the package is built around, so that every block has one
obvious place in a pipeline.

## Verbs compute, nouns render

The organizing principle is a split between two kinds of blocks:

- **Verbs compute** and return a data frame you can keep transforming. The
  one verb blockr.viz ships is `summary_table`; you compose it with dplyr verbs
  (`summarize`, `filter`, `pivot_wider`) from blockr.dplyr. (Statistical shapers
  — model fitting, `broom` adapters, and the like — live in sibling packages
  such as blockr.stats.)
- **Nouns render**: they take an already-shaped data frame and draw it
  (`chart`, `table`, `tile`, `gt`).

Two invariants make the system predictable:

1. If a block's name is a *verb* it returns a data frame. If it's a bare
   *noun* it draws.
2. **Renderers only ever see a rectangular data frame** (tidy or wide). The only non-data-frame
   intermediate in the whole system is a fitted statistical model, and it is
   fenced off — it exists solely between a `model` block and a `broom` adapter.
   No renderer ever touches a fit.

## The five layers

Every block in a dashboard pipeline is one of five kinds. Data flows left to
right, and a renderer only ever sees a rectangular data frame.

| Layer | Job | Output | Blocks |
|---|---|---|---|
| **Shape** | compute / summarise | tidy df | `summary_table` (blockr.viz) + dplyr verbs (`summarize`, `filter`, …); statistical shapers live in blockr.stats |
| **Reshape** | tidy → display grid (wide, *on purpose*) | wide df | dplyr `pivot_wider` |
| **Fit** | fit a statistical model | model object | `model` |
| **Adapt** | model → tidy | tidy df | `broom` (tidy / glance / augment) |
| **Render** | df → pixels | UI | `table`, `chart`, `tile`, `gt` |

So a pipeline reads as a sequence of roles:

- `summary_table → table` — summarise variables by a group, shown as a Table 1
- `summary_table → gt` — the same, rendered for print
- a correlation matrix `→ table` — a diverging, cell-colored heatmap
- `summarize → pivot_wider → chart` — a crosstab plotted
- `model → broom → table` — a coefficient table (`model`/`broom` from blockr.stats)

`blockr.viz` owns the **render** layer, plus the `summary_table` shaper that
feeds it.

## The render layer

### Interactive | static pairs

Each output medium has an **interactive** renderer (dashboard-native) and,
where it makes sense, a **static** renderer (print / report / CSR):

| Medium | Interactive | Static |
|---|---|---|
| table | `new_table_block()` | `new_gt_table_block()` |
| chart | `new_chart_block()` (ECharts) | `ggplot` (separate package) |
| number / spark / ring | `new_tile_block()` | — |

Pick by destination: dashboards use `table` / `chart` / `tile`; print and CSR
output use `gt`. The static renderers conceptually belong with their toolkit
packages; blockr.viz owns the interactive ones (and ships a convention-aware
`gt` adapter).

### Drill-down

Both `chart` and `table` can **drill**: clicking a mark or a row emits a filter
to downstream blocks. It is **opt-in and off by default** — one capability of the
general renderers, not a separate kind of block, so you reach for the same
`chart` and `table` for almost every chart or table in a dashboard.

To enable it, set the `drill` argument to a column name:

```{r}
library(blockr.viz)

# A bar chart; clicking a bar filters downstream on SITEID
new_chart_block(
  group = "SITEID", color = "TRT01P",
  chart_type = "bar", agg_fn = "count",
  drill = "SITEID"
)

# A table whose row click filters downstream on USUBJID
new_table_block(
  rowname = "USUBJID",
  values = c("AGE", "BMIBL", "WEIGHTBL"),
  drill = "USUBJID"
)
```

When `drill` is unset, a click is inert and no filter is emitted.

### `table` is the universal interactive renderer

`new_table_block()` handles both **flat** rectangular frames (with optional
cell coloring and drill) and **structured** "Table 1" input — nested row-side
section headers, multi-level column spanners, and collapse/expand — using the
same builders as the static `html_table()` path. One renderer covers flat and
structured tables alike. This reflects the package's strategy: deepen a few
general renderers rather than adding many specialized blocks.

### Renderers draw and format; they do not compute

A renderer **draws** structure (sections, spanners, indents), **formats**
numbers (decimals), and offers color, drill, and sort/search. It does **not**
aggregate, reshape, or run statistics.

The one bounded exception is a chart, which self-aggregates because the visual
encoding *is* the aggregation (a bar is a sum-over-group). Charts take raw data
plus an aesthetic mapping; tables take an already-shaped table.

## The `.fmt` cell convention

Composite cells like the clinical `"38 (54%)"` (count and percent) sit right on
the compute/draw boundary. blockr.viz resolves this with a hidden dotted column
called `.fmt`, a sibling of `.indent` / `.label` / `.section_*`. The raw numbers
stay in plain numeric columns (so the frame remains `dplyr`-able), and `.fmt`
carries a per-row **template** that names which columns to combine:

```
 .fmt = "{n} ({pct}%)"                  ->  "38 (54%)"
 .fmt = "{estimate} ({std.error})"      ->  "0.05 (0.01)"
 .fmt = "{median} ({q1}, {q3})"         ->  "12 (8, 19)"
```

The renderer is **pure interpolation**: it glues the named columns into the
literal template. There is no codebook of named formats and no domain
knowledge. Rounding stays out of the template — a `digits` hint owns it.

Because the numbers stay tidy, the *same* frame can render as combined cells
**or** as separate columns — a render-time choice, not something baked in by the
shaper. For grouped tables the cells are formatted **per row first**, then a
plain `pivot_wider()` spreads the grouping dimension to columns.

This is why one `table` renderer plus the `.fmt` convention covers Table 1,
regression (`broom`) output, survival tables, and frequency tables: they are all
the same display — rows by an optional spread dimension by composite cells — fed
by tidy shapers.

## The shaper: `summary_table()`

`summary_table()` (and its block `new_summary_table_block()`) is the one shaper
that lives in blockr.viz. It aggregates a flat data frame into a tidy, long
summary frame — raw numeric statistic columns plus the dotted structure columns
and a `.fmt` template — covering the "list of variables by Y" pattern:

```{r, eval = TRUE}
library(blockr.viz)

summary_table(
  iris,
  vars = c("Sepal.Length", "Species"),
  by = character(),
  stats = "compact"
)
```

Note the output: one `Mean (SD)` row for the numeric variable and one row per
level for the categorical, every statistic in a plain numeric column, with the
`.fmt` template carrying the display recipe. Its output is tidy and re-renderable by
any of the renderers above. The complementary "one measurement across two
dimensions" pivot is not a bespoke block — it is composed upstream from
`summarize` and `tidyr::pivot_wider`.

## Putting it together

Call `register_viz_blocks()` once to register the five blocks with the
block-adder and the assistant. See the package README for a runnable end-to-end
board.
