---
title: "Shapers and renderers: how blockr.viz is organized"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shapers and renderers: how blockr.viz is organized}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

blockr.viz is the render layer for blockr dashboards. It draws rectangular data
frames; it does not compute or reshape them. This vignette describes the block
roles the package uses, so each block has one place in a pipeline.

## Verbs and nouns

Blocks come in two kinds.

- Verb blocks compute and return a data frame you can keep transforming.
  blockr.viz ships one, `summary_table`; you compose it with the dplyr verbs
  (`summarize`, `filter`, `pivot_wider`) from blockr.dplyr. Model fitting and
  `broom` adapters are in blockr.stats.
- Noun blocks render: they take a data frame and draw it (`chart`, `table`,
  `tile`, `gt`).

Two rules follow. A verb-named block returns a data frame; a noun-named block
draws. And a renderer only ever receives a rectangular data frame, tidy or wide.
The one non-data-frame value in the system is a fitted model, which passes only
from a `model` block to a `broom` adapter; no renderer sees a model.

## The five layers

A dashboard pipeline is built from five kinds of block. Data flows left to
right, and a renderer receives a rectangular data frame.

| Layer | Job | Output | Blocks |
|---|---|---|---|
| Shape | compute, summarise | tidy df | `summary_table` (blockr.viz), dplyr verbs (`summarize`, `filter`, …); statistical shapers in blockr.stats |
| Reshape | tidy to a wide display grid | wide df | dplyr `pivot_wider` |
| Fit | fit a statistical model | model object | `model` |
| Adapt | model to tidy df | tidy df | `broom` (tidy, glance, augment) |
| Render | df to pixels | UI | `table`, `chart`, `tile`, `gt` |

Example pipelines:

- `summary_table → table`: summarise variables by a group, shown as a Table 1
- `summary_table → gt`: the same table, rendered for print
- correlation matrix `→ table`: a diverging, cell-colored heatmap
- `summarize → pivot_wider → chart`: a crosstab plotted
- `model → broom → table`: a coefficient table (`model` and `broom` from blockr.stats)

blockr.viz provides the render layer and the `summary_table` shaper.

## The render layer

### Interactive and static renderers

Each medium has an interactive renderer for dashboards and, where it applies, a
static renderer for print or report output.

| Medium | Interactive | Static |
|---|---|---|
| table | `new_table_block()` | `new_gt_table_block()` |
| chart | `new_chart_block()` (ECharts) | `ggplot` (separate package) |
| number / spark / ring | `new_tile_block()` | — |

Dashboards use `table`, `chart`, and `tile`; print output uses `gt`. blockr.viz
provides the interactive renderers and a convention-aware `gt` adapter; the
`ggplot` static renderer lives with its own package.

### Drill-down

`chart` and `table` can drill: clicking a mark or a row emits a filter to
downstream blocks. Drill is opt-in and off by default. It is one argument on the
general renderers, not a separate block. Set `drill` to a column name to enable
it.

```{r}
library(blockr.viz)

# A bar chart; clicking a bar filters downstream on SITEID
new_chart_block(
  group = "SITEID", color = "TRT01P",
  chart_type = "bar", agg_fn = "count",
  drill = "SITEID"
)

# A table whose row click filters downstream on USUBJID
new_table_block(
  rowname = "USUBJID",
  values = c("AGE", "BMIBL", "WEIGHTBL"),
  drill = "USUBJID"
)
```

With `drill` unset, a click emits nothing.

### One table renderer for flat and structured input

`new_table_block()` renders flat rectangular frames, with optional cell coloring
and drill, and structured "Table 1" input: nested row-side section headers,
multi-level column spanners, and collapse/expand. Both use the same builders as
the static `html_table()` path, so one renderer covers flat and structured
tables.

### What a renderer does and does not do

A renderer draws structure (sections, spanners, indents), formats numbers, and
offers color, drill, and sort/search. It does not aggregate, reshape, or run
statistics.

A chart is the exception: it self-aggregates, because the visual encoding is the
aggregation (a bar is a sum over a group). Charts take raw data and an aesthetic
mapping; tables take a shaped table.

## The `.fmt` cell convention

A composite cell such as `"38 (54%)"` (count and percent) sits on the
compute/draw boundary. blockr.viz keeps the numbers in plain numeric columns and
stores the display recipe in a hidden column, `.fmt`, alongside `.indent`,
`.label`, and `.section_*`. `.fmt` holds a per-row template naming the columns to
combine:

```
 .fmt = "{n} ({pct}%)"                  ->  "38 (54%)"
 .fmt = "{estimate} ({std.error})"      ->  "0.05 (0.01)"
 .fmt = "{median} ({q1}, {q3})"         ->  "12 (8, 19)"
```

The renderer interpolates: it substitutes the named columns into the template.
There is no table of named formats. Rounding is not part of the template; a
`digits` hint controls it.

Because the numbers stay numeric, the same frame can render as combined cells or
as separate columns, a choice made at render time. For grouped tables the cells
are formatted per row, then `pivot_wider()` spreads the grouping dimension to
columns.

One `table` renderer with this convention covers Table 1, regression output
(`broom`), survival tables, and frequency tables. Each is the same display: rows,
an optional spread dimension, and composite cells, fed by a shaper.

## The `summary_table` shaper

`summary_table()` and its block `new_summary_table_block()` are the shaper in
blockr.viz. It aggregates a flat data frame into a long summary frame: raw
numeric statistic columns, the dotted structure columns, and a `.fmt` template.
It covers the "list of variables by group" pattern.

```{r, eval = TRUE}
library(blockr.viz)

summary_table(
  iris,
  vars = c("Sepal.Length", "Species"),
  by = character(),
  stats = "compact"
)
```

The output has one `Mean (SD)` row for the numeric variable and one row per level
for the categorical. Every statistic is a plain numeric column, and `.fmt`
carries the display recipe, so any renderer above can consume it. The "one
measurement across two dimensions" pivot is not a separate block; compose it from
`summarize` and `tidyr::pivot_wider`.

## Registering the blocks

Call `register_viz_blocks()` once to register the five blocks with the
block-adder and the assistant. The package README has a runnable end-to-end
board.
